---
title: "BIS15L Project"
output: 
  html_document: 
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##MRI Data and Alzhiemer's Disease

In this project, we are using data from the Open Access Series of Imaging Studies (OASIS), made available by the Washingtom UNiversity Alzheimer's Disease Research Center, Dr. Randy Buckner at the Howard Hughes Medical Institute (HHMI), the Neuroinformatics Research Group (NRG) at Washington University School of Medicine, and the Biomedical Informatics Research Network (BIRN). [MRI and Alzheimer's Disease](https://www.kaggle.com/jboysen/mri-and-alzheimers)

Load libraries
```{r}
library(tidyverse)
library(RColorBrewer)
library(paletteer)
library(janitor)
library(here)
library("RColorBrewer")
library("paletteer")
library(viridis)
library(GGally)
```

```{r}
options(scipen=999)
```

```{r}
longitudinal <- readr::read_csv("data/oasis_longitudinal.csv")
xsectional <- readr::read_csv("data/oasis_cross-sectional.csv")
```
```{r}
glimpse(longitudinal)
```

```{r}
summary(longitudinal)
```
```{r}
longitudinal <- janitor::clean_names(longitudinal)
xsectional <- janitor::clean_names(xsectional)
```

```{r}
names(longitudinal)
names(xsectional)
```
```{r}
longitudinal$age <- as.factor(longitudinal$age)
```


```{r}
longitudinal %>% 
  ggplot(aes(x=age, y=n_wbv, fill=m_f))+
  geom_col()
```

```{r}
longitudinal %>% 
  ggplot(aes(x=educ))+geom_bar()
```
```{r}
names(longitudinal)
```

**not sure why scale_fill_brewer isnt working**
```{r}
age_xsectional <- xsectional %>% 
  filter(cdr!='NA') %>% 
  ggplot(aes(x=age))+
  geom_density(alpha=0.5)
age_xsectional+facet_wrap(~cdr) +
  theme(axis.text.x = element_text(angle=60, hjust=1), plot.title = element_text(size = rel(1), hjust = 0.5))+
  scale_fill_brewer(palette = "Set1")+
  labs(title = "Age Facet Wrap", x= "Age")
?scale_fill_brewer 
```
```{r}
asf_long <- longitudinal %>% 
  filter(age!='NA' | asf!='NA') %>% 
  group_by(subject_id) %>% 
  ggplot(aes(x=age, y=asf, fill=group))+
  geom_col(alpha=0.5)+
  theme(axis.text.x = element_text(angle = 60, hjust = 1, size = 8), plot.title = element_text(size = rel(1), hjust = 0.5))+
  labs(title = "Distribution of Atlas Scaling Factor", x= "Age", y="Atlas Scaling Factor")
asf_long
```
```{r}
asf_xsectional <- xsectional %>% 
  filter(age!='NA' | asf!='NA') %>% 
  ggplot(aes(x=age, y=asf))+
  geom_col(alpha=0.5)+
  theme(axis.text.x = element_text(angle = 60, hjust = 1, size = 8), plot.title = element_text(size = rel(1), hjust = 0.5))+
  labs(title = "Distribution of Atlas Scaling Factor", x= "Age", y="Atlas Scaling Factor")
asf_xsectional
```

```{r}
age_xsectional <- xsectional %>% 
  filter(educ!='NA') %>% 
  ggplot(aes(x=age))+
  geom_density(alpha=0.5)
age_xsectional+facet_wrap(~educ) +
  theme(axis.text.x = element_text(angle=60, hjust=1), plot.title = element_text(size = rel(1), hjust = 0.5))+
  scale_fill_brewer(palette = "Set1")+
  labs(title = "EDUC Facet Wrap", x= "Age")
```

```{r}
educ_xsectional <- xsectional %>% 
  filter(age!='NA' | educ!='NA') %>% 
  ggplot(aes(x=age, y=educ))+
  geom_col(alpha=0.5)+
  theme(axis.text.x = element_text(angle = 60, hjust = 1, size = 8), plot.title = element_text(size = rel(1), hjust = 0.5))+
  #scale_x_discrete(breaks=c(20, 30, 40, 50, 60, 70, 80, 90))+
  labs(title = "Distribution of Education Status", x= "Age", y="Education Status")
educ_xsectional
```
```{r}
educ_long <- longitudinal %>% 
  filter(age!='NA' | educ!='NA') %>% 
  ggplot(aes(x=age, y=educ, fill=group))+
  geom_col(alpha=0.5)+
  theme(axis.text.x = element_text(angle = 60, hjust = 1, size = 8), plot.title = element_text(size = rel(1), hjust = 0.5))+
  #scale_x_discrete(breaks=c(20, 30, 40, 50, 60, 70, 80, 90))+
  labs(title = "Distribution of Education Status", x= "Age", y="Education Status")
educ_long
```


```{r}
n_wbv_xsectional <- xsectional %>% 
  filter(age!='NA' | n_wbv!='NA') %>% 
  ggplot(aes(x=age, y=n_wbv, fill = n_wbv))+
  scale_fill_viridis(option = "D")+
  geom_col(alpha=0.7)+
  theme(axis.text.x = element_text(angle = 60, hjust = 1, size = 8), plot.title = element_text(size = rel(1), hjust = 0.5))+
  #scale_x_discrete(breaks=c(20, 30, 40, 50, 60, 70, 80, 90))+
  labs(title = "Distribution of Normalized Whole Brain Volume", x= "Age", y="Normalized Whole Brain Volume")
n_wbv_xsectional
```
```{r}
n_wbv_long <- longitudinal %>% 
  filter(age!='NA' | n_wbv!='NA') %>% 
  ggplot(aes(x=age, y=n_wbv, fill=asf))+
  geom_col(alpha=0.5)+
  scale_fill_viridis(option = "D")+
  theme(axis.text.x = element_text(angle = 60, hjust = 1, size = 8), plot.title = element_text(size = rel(1), hjust = 0.5))+
  #scale_x_discrete(breaks=c(20, 30, 40, 50, 60, 70, 80, 90))+
  labs(title = "Distribution of Normalized Whole Brain Volume", x= "Age", y="Normalized Whole Brain Volume")
n_wbv_long
```

```{r}
e_tiv_xsectional <- xsectional %>% 
  filter(age!='NA' | e_tiv!='NA') %>% 
  ggplot(aes(x=age, y=e_tiv))+
  geom_col(alpha=0.5)+
  theme(axis.text.x = element_text(angle = 60, hjust = 1, size = 8), plot.title = element_text(size = rel(1), hjust = 0.5))+
  #scale_x_discrete(breaks=c(20, 30, 40, 50, 60, 70, 80, 90))+
  labs(title = "Distribution of Estimated Total Intracranial Volume", x= "Age", y="Estimated Total Intracranial Volume")
e_tiv_xsectional
```
```{r}
e_tiv_long <- longitudinal %>% 
  filter(age!='NA' | e_tiv!='NA') %>% 
  ggplot(aes(x=age, y=e_tiv, fill=group))+
  geom_col(alpha=0.5)+
  scale_fill_viridis(discrete = TRUE, option = "B")+
  theme(axis.text.x = element_text(angle = 60, hjust = 1, size = 8), plot.title = element_text(size = rel(1), hjust = 0.5))+
  labs(title = "Distribution of Estimated Total Intracranial Volume", x= "Age", y="Estimated Total Intracranial Volume")
e_tiv_long
```
```{r}
ses_xsectional <- xsectional %>% 
  filter(age!='NA' | ses!='NA') %>% 
  ggplot(aes(x=age, y=ses, color = educ))+
  geom_point(alpha=0.5)+
  scale_color_viridis(option = "D")+
  theme(axis.text.x = element_text(angle = 60, hjust = 1, size = 8), plot.title = element_text(size = rel(1), hjust = 0.5))+
  #scale_x_discrete(breaks=c(20, 30, 40, 50, 60, 70, 80, 90))+
  labs(title = "Distribution of Socioeconomic Status", x= "Age", y="Socioeconomic Status")
ses_xsectional
```
```{r}
ses_xsectional <- xsectional %>% 
  filter(age!='NA' | ses!='NA') %>% 
  ggplot(aes(x=ses, y=educ, color = educ))+
  geom_point(alpha=0.5)+
  scale_color_viridis(option = "D")+
  theme(axis.text.x = element_text(angle = 60, hjust = 1, size = 8), plot.title = element_text(size = rel(1), hjust = 0.5))+
  #scale_x_discrete(breaks=c(20, 30, 40, 50, 60, 70, 80, 90))+
  labs(title = "Distribution of Socioeconomic Status", x= "EDUC", y="Socioeconomic Status")
ses_xsectional
```


```{r}
ses_long <- longitudinal %>% 
  filter(age!='NA' | ses!='NA') %>% 
  ggplot(aes(x=age, y=ses, fill=group))+
  geom_col(alpha=0.5)+
  theme(axis.text.x = element_text(angle = 60, hjust = 1, size = 8), plot.title = element_text(size = rel(1), hjust = 0.5))+
  labs(title = "Distribution of Socioeconomic Status", x= "Age", y="Socioeconomic Status")
ses_long
```

```{r}
asf_xsectional <- xsectional %>% 
  filter(age!='NA' | asf!='NA') %>% 
  ggplot(aes(x=age, y=asf))+
  geom_col(alpha=0.5)+
  theme(axis.text.x = element_text(angle = 60, hjust = 1, size = 8), plot.title = element_text(size = rel(1), hjust = 0.5))+
  labs(title = "Distribution of Atlas Scaling Factor", x= "Age", y="Atlas Scaling Factor")
asf_xsectional
```





```{r}
asf_long <- longitudinal %>% 
  filter(age!='NA' | asf!='NA') %>% 
  ggplot(aes(x=age, y=asf, color=group))+
  geom_point(alpha=0.5)+
  geom_smooth(method = "lm", se = FALSE)+
  scale_color_viridis(option = "D", discrete = TRUE)+
  theme(axis.text.x = element_text(angle = 60, hjust = 1, size = 8), plot.title = element_text(size = rel(1), hjust = 0.5))+
  labs(title = "Distribution of Atlas Scaling Factor", x= "Age", y="Atlas Scaling Factor")
asf_long
```
```{r}
mmse_xsectional <- xsectional %>% 
  filter(age!='NA' | mmse!='NA') %>% 
  group_by(age) %>% 
  mutate(mean_mmse = mean(mmse)) %>% 
  ggplot(aes(x=age, y=mmse))+
  geom_point(alpha=0.5)+
  geom_smooth(method = "lm", se = FALSE)+
  theme(axis.text.x = element_text(angle = 60, hjust = 1, size = 8), plot.title = element_text(size = rel(1), hjust = 0.5))+
  labs(title = "Distribution of Mini-Mental State Examination", x= "Age", y="Mini-Mental State Examination")
mmse_xsectional
```
```{r}
head(mmse_xsectional)
```


```{r}
mmse_long <- longitudinal %>% 
  group_by(subject_id) %>% 
  filter(age!='NA' | mmse!='NA') %>% 
  ggplot(aes(x=age, y=mmse, color=group))+
  geom_point(alpha=0.5)+
  geom_smooth(method = "lm", se = FALSE)+
  scale_color_viridis(option = "D", discrete = TRUE)+
  theme(axis.text.x = element_text(angle = 60, hjust = 1, size = 8), plot.title = element_text(size = rel(1), hjust = 0.5))+
  labs(title = "Distribution of Mini-Mental State Examination", x= "Age", y="Mini-Mental State Examination")
mmse_long
```


```{r}
age_long <- longitudinal %>% 
  filter(age!='NA' | group!='NA') %>%  
  filter(group=="Demented") %>% 
  ggplot(aes(x=age, fill=group))+
  geom_bar(alpha=0.5)+
  theme(axis.text.x = element_text(angle = 60, hjust = 1, size = 8), plot.title = element_text(size = rel(1), hjust = 0.5))+
  labs(title = "Distribution of Group", x= "Age")
age_long
```
```{r}
age_long <- longitudinal %>% 
  filter(age!='NA' | group!='NA') %>%  
  filter(group=="Nondemented") %>% 
  ggplot(aes(x=age, fill=group))+
  geom_bar(alpha=0.5)+
  theme(axis.text.x = element_text(angle = 60, hjust = 1, size = 8), plot.title = element_text(size = rel(1), hjust = 0.5))+
  labs(title = "Distribution of Group", x= "Age")
age_long
```
```{r}
age_long <- longitudinal %>% 
  filter(age!='NA' | group!='NA') %>%  
  filter(group=="Converted") %>% 
  ggplot(aes(x=age, fill=group))+
  geom_bar(alpha=0.5)+
  theme(axis.text.x = element_text(angle = 60, hjust = 1, size = 8), plot.title = element_text(size = rel(1), hjust = 0.5))+
  labs(title = "Distribution of Group", x= "Age")
age_long
```
```{r}
age_long <- longitudinal %>% 
  filter(age!='NA' | group!='NA') %>%  
  filter(group=="Converted") %>% 
  ggplot(aes(x=age, fill=group))+
  geom_bar(alpha=0.5)+
  theme(axis.text.x = element_text(angle = 60, hjust = 1, size = 8), plot.title = element_text(size = rel(1), hjust = 0.5))+
  labs(title = "Distribution of Group", x= "Age")
age_long
```
```{r}

```



COUNT Graphs
**AGE**
```{r}
xsectional %>% 
  ggplot(aes(x=age, fill= "red", alpha= 0.5))+
  geom_bar()+
  labs(title="Cross-Sectional Age Distribution",
       x=NULL,
       y="Count")+
  theme_classic()+
  theme(plot.title = element_text(hjust = .5), legend.position = "none")

longitudinal %>% 
  group_by("subject_id") %>% 
  ggplot(aes(x=age, fill= age, alpha = 0.5))+
  geom_bar()+
  labs(title="Longitudinal Age Distribution",
       x=NULL,
       y="Count")+
  theme_classic()+
  theme(plot.title = element_text(hjust = .5),legend.position = "none")
```
**SES**
```{r}
xsectional %>% 
  ggplot(aes(x=ses, fill="blue", alpha = 0.7))+
  geom_bar()+
  labs(title="Cross-Sectional Socioeconomic Status Distribution",
       x=NULL,
       y="Count")+
  theme_classic()+
  theme(plot.title = element_text(hjust = .5), legend.position = "none")

longitudinal %>% 
  group_by("subject_id") %>% 
  ggplot(aes(x=as.factor(ses), fill=as.factor(ses), alpha = 0.7))+
  geom_bar()+
  labs(title="Longitudinal Socioeconomic Status Distribution",
       x=NULL,
       y="Count")+
  theme_classic()+
  theme(plot.title = element_text(hjust = .5), legend.position = "none")
```

**Sex**
```{r}
xsectional %>% 
  ggplot(aes(x=m_f, fill = "red", alpha = 0.5))+
  geom_bar()+
  labs(title="Cross-Sectional Sex Distribution",
       x=NULL,
       y="Count")+
  theme_classic()+
  theme(plot.title = element_text(hjust = .5), legend.position = "none")

longitudinal %>% 
  group_by("subject_id") %>% 
  ggplot(aes(x=m_f, fill = m_f, alpha = 0.5))+
  geom_bar()+
  labs(title="Longitudinal Sex Distribution",
       x=NULL,
       y="Count")+
  theme_classic()+
  theme(plot.title = element_text(hjust = .5), legend.position = "none")
```
**Education**
```{r}
xsectional %>% 
  ggplot(aes(x=educ, fill = "red", alpha = 0.5))+
  geom_bar()+
   labs(title="Cross-Sectional Education Status Distribution",
       x=NULL,
       y="Count")+
  theme_classic()+
  theme(plot.title = element_text(hjust = .5), legend.position = "none")

longitudinal %>% 
  group_by("subject_id") %>% 
  ggplot(aes(x=educ, fill = , alpha=0.5))+
  geom_bar()+
   labs(title="Longitudinal Education Status Distribution",
       x=NULL,
       y="Count")+
  theme_classic()+
  theme(plot.title = element_text(hjust = .5), legend.position = "none")
```
```{r}

longitudinal %>% 
  group_by("subject_id") %>% 
  ggplot(aes(x=group, fill = group, alpha=0.5))+
  geom_bar()+
  scale_color_viridis(option = "B")+
   labs(title="Longitudinal Dementia Status Distribution",
       x=NULL,
       y="Count")+
  theme_classic()+
  theme(plot.title = element_text(hjust = .5), legend.position = "none")
```



```{r}
#install.packages("ggcorrplot")
library(ggcorrplot)
library(ggplot2)
#install.packages("GGally")
library("GGally")
```
```{r}
#install.packages("viridis")
library(viridis)
```

**Correlation plot**
```{r}
ggcorr(longitudinal, label = TRUE, label_size = 3, size = 4, hjust = 0.75)+
  labs(title="Longitudinal Correlation Plot")+
  theme(plot.title = element_text(hjust = 0.5))
  



ggcorr(xsectional, label = TRUE, label_size = 3)+
  labs(title="Cross-Sectional Correlation Plot")+
  theme(plot.title = element_text(hjust = .5))
```
```{r}
?ggcorr
```

```{r}
long_cor <- longitudinal %>% 
  select(age, educ, ses, mmse, cdr, e_tiv, n_wbv, asf) %>% 
  na.exclude
long_cor <- round(cor(long_cor), 2)
head(long_cor)
```
```{r}
library(reshape2)
melted_long <- melt(long_cor)
head(melted_long)
ggplot(data = melted_long, aes(x=Var1, y=Var2, fill=value)) + 
  geom_tile(label = TRUE)
```
```{r}
l_cor <- longitudinal %>% 
  select(age, educ, ses, mmse, cdr, e_tiv, n_wbv, asf) %>% 
  na.exclude
ggpairs(l_cor, title="correlogram with ggpairs()", size = 10) 
```

```{r}
?ggpairs
```

```{r}
xsec_cor <- xsectional %>% 
  select(age, educ, ses, mmse, cdr, e_tiv, n_wbv, asf) %>% 
  na.exclude
ggpairs(xsec_cor, title="correlogram with ggpairs()") 
```

```{r}
df_cross <- xsectional
df_long <- longitudinal
glimpse(df_cross)
glimpse(df_long)
```
```{r}
df_cross <- df_cross %>%
  select(-c("id", "hand", "delay"))
#colnames(df_cross)[3] <- "educ"
df_cross <- df_cross[complete.cases(df_cross),]
df_long <- df_long %>%
  select(-c(subject_id, mri_id, hand, visit, mr_delay, group))
df <- rbind(df_cross,df_long)
glimpse(df)
```