---
title: "BIS15L Final Project Group 9"
output: 
  html_document: 
    keep_md: yes
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# BIS15L Final Project Group 9

In this project, we are using data from the Open Access Series of Imaging Studies (OASIS), made available by the Washingtom UNiversity Alzheimer's Disease Research Center, Dr. Randy Buckner at the Howard Hughes Medical Institute (HHMI), the Neuroinformatics Research Group (NRG) at Washington University School of Medicine, and the Biomedical Informatics Research Network (BIRN). Our datasets were found on kaggle: [MRI and Alzheimer's Disease](https://www.kaggle.com/jboysen/mri-and-alzheimers).

The question we wanted to answer is "What are the differences in measured variables between Demented and Non-demented subjects and do they have any predictive value in the development of Alzheimer's Disease Dementia?"


Load libraries

```{r}
library(tidyverse)
library(RColorBrewer)
library(paletteer)
library(janitor)
library(here)
library(GGally)
library(shiny)
library(shinydashboard)
library(viridis)
```

```{r}
options(scipen=999)
```

```{r}
longitudinal <- readr::read_csv("data/oasis_longitudinal.csv")
xsectional <- readr::read_csv("data/oasis_cross-sectional.csv")
```

**Let's take a look at the data: **


```{r}
glimpse(longitudinal)
summary(longitudinal)
```

```{r}
glimpse(xsectional)
summary(xsectional)
```

Clean Names

```{r}
longitudinal <- janitor::clean_names(longitudinal)
xsectional <- janitor::clean_names(xsectional)
```

```{r}
names(longitudinal)
names(xsectional)
```
# Explanation of Datasets

The **Longitudinal Dataset** includes data from 150 patients, aged 60-96 years old. Each subject received an MRI scan on two or more visits, separated by at least one year for a total of 373 imaging sessions. In this dataset, 72 patients were characterized as Nondemented, 64 were characterized as demented (with 51 subjects having mild to moderate dementia), and 14 were converted throughout the study.

The **Cross Sectional Dataset** includes data from 416 subjects aged 18 to 96 years old. In this dataset, 100 subjects over the age of 60 were clinically diagnosed with very mild to moderate Alzheimer’s disease, and 20 nondemented subjects were used as a control in the study.


# Explanation of Variables
**CDR** - Clinical Dementia Rating - Used to group patients based on the severity of dementia (0 = no dementia, 0.5 = very mild AD, 1 = mild AD, 2 = moderate AD)

**eTIV** - Estimated Total Intracranial Volume 

**nWBV** - Normalized Whole Brain Volume - as measured by mRI

**ASF** - Atlas Scaling Factor - factor that allows for comparison of estimated total intracranial volume based on differences in human anatomy

**MR Delay** - Time between appointments

**Group** - demented/nondemented/converted (Coverted means that they did not have dementia at their first visit to the doctor, but was diagnosed with dementia at a subsequent visit during the study).

**Hand** - All patients are right handed

**Age** - Age in years 

**EDUC** - Education Status (measured in years)

**SES** - socioeconomic status as measured by the Hollingshead Index of Social Position, where 1 = highest status and 5 = lowest status



# Cursory Analysis of the data: 


### Age Distribution Plots of Both Datasets
```{r}
xsectional %>% 
  ggplot(aes(x=age, fill= as.factor(age), alpha= 0.5))+
  geom_bar()+
  labs(title="Cross-Sectional Age Distribution",
       x="Age",
       y="Count")+
  scale_fill_viridis(option = "B", discrete = TRUE)+
  theme_classic()+
  theme(axis.text.x = element_text(angle = 60, hjust = 1, size = 7, ), plot.title = element_text(hjust = .5), legend.position = "none")

longitudinal %>% 
  group_by("subject_id") %>% 
  ggplot(aes(x=age, fill= as.factor(age), alpha = 0.5))+
  geom_bar()+
  labs(title="Longitudinal Age Distribution",
       x="Age",
       y="Count")+
  scale_fill_viridis(option = "B", discrete = TRUE)+
  theme_classic()+
  theme(axis.text.x = element_text(angle = 60, hjust = 1, size = 8), plot.title = element_text(hjust = .5),legend.position = "none")
```

### Socioeconomic Status Distribution Plots of Both Datasets

```{r}
xsectional %>% 
  ggplot(aes(x=ses, fill=as.factor(ses), alpha = 0.7))+
  geom_bar()+
  labs(title="Cross-Sectional Socioeconomic Status Distribution",
       x="Socioeconomic Status",
       y="Count")+
  theme_classic()+
  scale_fill_viridis(option = "B", discrete = TRUE)+
  theme(plot.title = element_text(hjust = .5), legend.position = "none")

longitudinal %>% 
  group_by("subject_id") %>% 
  filter(ses != "NA") %>% 
  ggplot(aes(x=as.factor(ses), fill=as.factor(ses), alpha = 0.7))+
  geom_bar()+
  labs(title="Longitudinal Socioeconomic Status Distribution",
       x="Socioeconomic Status",
       y="Count")+
  theme_classic()+
  scale_fill_viridis(option = "B", discrete = TRUE)+
  theme(plot.title = element_text(hjust = .5), legend.position = "none")
```


### Sex Distribution Plots of Both Datasets

```{r}
xsectional %>%
  ggplot(aes(x=m_f, fill = m_f, alpha = 0.5))+
  geom_bar()+
  labs(title="Cross-Sectional Sex Distribution",
       x="Sex",
       y="Count")+
  theme_classic()+
  scale_fill_viridis(option = "B", discrete = TRUE)+
  theme(plot.title = element_text(hjust = .5), legend.position = "none")

longitudinal %>% 
  group_by("subject_id") %>% 
  ggplot(aes(x=m_f, fill = m_f, alpha = 0.5))+
  geom_bar()+
  labs(title="Longitudinal Sex Distribution",
       x="Sex",
       y="Count")+
  theme_classic()+
  scale_fill_viridis(option = "B", discrete = TRUE)+
  theme(plot.title = element_text(hjust = .5), legend.position = "none")
```

### Education Status Distribution Plots of Both Datasets

```{r}
xsectional %>% 
  ggplot(aes(x=educ, fill = as.factor(educ), alpha = 0.5))+
  geom_bar()+
   labs(title="Cross-Sectional Education Status Distribution",
       x="Education Status",
       y="Count")+
  theme_classic()+
  scale_fill_viridis(option = "B", discrete = TRUE)+
  theme(plot.title = element_text(hjust = .5), legend.position = "none")

longitudinal %>% 
  group_by("subject_id") %>% 
  ggplot(aes(x=educ, fill = as.factor(educ), alpha=0.5))+
  geom_bar()+
   labs(title="Longitudinal Education Status Distribution",
       x="Education Status",
       y="Count")+
  theme_classic()+
  scale_fill_viridis(option = "B", discrete = TRUE)+
  theme(plot.title = element_text(hjust = .5), legend.position = "none")
```

### Group Distribution Plot of Longitudinal Dataset

```{r}
longitudinal %>% 
  group_by("subject_id") %>% 
  ggplot(aes(x=group, fill = group, alpha=0.5))+
  geom_bar()+
   labs(title="Longitudinal Dementia Status Distribution",
       x=NULL,
       y="Count")+
  theme_classic()+
  scale_fill_viridis(option = "C", discrete = TRUE)+
  theme(plot.title = element_text(hjust = .5), legend.position = "none")
```

### Correlation Plots of Both Datasets

```{r}
ggcorr(longitudinal, label = TRUE, label_size = 3, size = 4, hjust = 0.75)+
  labs(title="Longitudinal Correlation Plot")+
  theme(plot.title = element_text(hjust = 0.5))

  


ggcorr(xsectional, label = TRUE, label_size = 3)+
  labs(title="Cross-Sectional Correlation Plot")+
  theme(plot.title = element_text(hjust = .5))
```

### How does normalized whole brain volume and CDR change with age?

```{r}
n_wbv_xsectional <- xsectional %>% 
  filter(age!='NA' | n_wbv!='NA') %>% 
  ggplot(aes(x=age, y=n_wbv, color = cdr))+
  scale_color_viridis(option = "B")+
  geom_smooth(method = "lm")+
  geom_point()+
  theme_classic()+
  theme(axis.text.x = element_text(angle = 60, hjust = 1, size = 8), plot.title = element_text(size = rel(1), hjust = 0.5))+
  labs(title = "Distribution of Normalized Whole Brain Volume", x= "Age", y="Normalized Whole Brain Volume")
n_wbv_xsectional
```

### What is the relationship between CDR and Normalized Whole Brain Volume?


```{r}
xsectional %>% 
  group_by(as.factor(cdr)) %>% 
  ggplot(aes(x=as.factor(cdr),y=n_wbv,group=as.factor(cdr),color=as.factor(cdr),fill=as.factor(cdr),alpha=.5))+
  geom_boxplot()+
  geom_jitter()+
  theme_classic()+
  theme(plot.title = element_text(hjust = .5),legend.position = "none")+
  labs(title=" Normalized Whole Brain Volume vs Clinical Dementia Rating",
       x="Clinical Dementia Rating",
       y="Normalized Whole Brain Volume")+
  scale_x_discrete(labels=c("NA" = "Young"))
```

### What is the Distribution of Age within CDR groups?


```{r}
xsectional %>% 
  filter(cdr!="NA", age>60) %>% 
  ggplot(aes(x = as.factor(cdr), y = age, group=cdr,color=as.factor(cdr))) + 
  geom_violin(aes(fill = as.factor(cdr),alpha=.5)) + 
  geom_boxplot(width = 0.2, alpha=1)+
  theme_classic()+
  labs(title="Age Distribution Between Groups",
       x="CDR",
       y="Age Distribution ")+
  theme(plot.title = element_text(hjust = .5),legend.position = "none")
```

### Years of Education Between Groups

```{r}
longitudinal %>%
  filter(age>60,group!="Converted",visit==1) %>% 
  group_by(group) %>% 
  ggplot(aes(x=group,y=educ,fill=group,color=group,alpha=.5))+
  geom_boxplot()+
  theme_classic()+
  labs(title="Years of Education vs Dementia Status",
       x=NULL,
       y="Education")+
  theme(plot.title = element_text(hjust = .5),legend.position = "none")+
  scale_x_discrete(labels=c("Nondemented" = "Non-Demented"))
```

### SES Distribution Between Groups

```{r}
longitudinal %>% 
  filter(age>60,group!="Converted",ses!="NA",visit==1) %>% 
  group_by(group) %>% 
  ggplot(aes(x=group,y=ses,fill=group,color=group,alpha=.5))+
  geom_boxplot()+
  labs(title="SES vs Dementia Status",
       x=NULL,
       y="SES")+
  theme_classic()+
  theme(plot.title = element_text(hjust = .5),legend.position = "none")+
  scale_x_discrete(labels=c("Nondemented" = "Non-Demented"))
```

### How does normalized whole brain volume change over subsequent visits?

```{r}
longitudinal %>% 
  filter(group=="Nondemented") %>% 
  ggplot(aes(x=visit,y=n_wbv,color=subject_id))+ 
  geom_point()+
  geom_line()+
  theme_classic()+
  labs(title="Change in Normalized Whole Brain Volume Over Time (Non-Demented|CDR=0)",
       x="Visit Number",
       y="Normalized Whole Brain Volume")+
  theme(plot.title = element_text(hjust = .5),legend.position = "none")
```
```{r}
longitudinal %>% 
  filter(group=="Demented") %>% 
  ggplot(aes(x=visit,y=n_wbv,color=subject_id))+ 
  geom_point()+
  geom_line()+
  theme_classic()+
  labs(title="Change in Normalized Whole Brain Volume Over Time (Demented|CDR>0)",
       x="Visit Number",
       y="Normalized Whole Brain Volume")+
  theme(plot.title = element_text(hjust = .5),legend.position = "none")
```
```{r}
longitudinal %>% 
  filter(group=="Converted") %>% 
  ggplot(aes(x=visit,y=n_wbv,color=subject_id))+ 
  geom_point()+
  geom_line()+
  theme_classic()+
  labs(title="Change in Normalized Whole Brain Volume Over Time (Converted)",
       x="Visit Number",
       y="Normalized Whole Brain Volume")+
  theme(plot.title = element_text(hjust = .5),legend.position = "none")
```
```{r}
longitudinal %>% 
  filter(group=="Nondemented") %>% 
  ggplot(aes(x=visit,y=n_wbv,group=group,color=subject_id))+ 
  geom_point()+
  geom_smooth(method="lm",se=TRUE)+
  theme_classic()+
  labs(title="Change in Normalized Whole Brain Volume Over Time (Nondemented)",
       x="Visit Number",
       y="Normalized Whole Brain Volume")+
  theme(plot.title = element_text(hjust = .5),legend.position = "none")
```
```{r}
longitudinal %>% 
  filter(group=="Demented") %>% 
  ggplot(aes(x=visit,y=n_wbv,group=group,color=subject_id))+ 
  geom_point()+
  geom_smooth(method=lm,se=TRUE,fullrange=TRUE)+
  theme_classic()+
  labs(title="Change in Normalized Whole Brain Volume Over Time (Demented)",
       x="Visit Number",
       y="Normalized Whole Brain Volume")+
  theme(plot.title = element_text(hjust = .5),legend.position = "none")
```
```{r}
longitudinal %>% 
  filter(group=="Converted") %>% 
  ggplot(aes(x=visit,y=n_wbv,group=group,color=subject_id))+ 
  geom_point()+
  geom_smooth(method=lm,se=TRUE,fullrange=TRUE)+
  theme_classic()+
  labs(title="Change in Normalized Whole Brain Volume Over Time (Converted)",
       x="Visit Number",
       y="Normalized Whole Brain Volume")+
  theme(plot.title = element_text(hjust = .5),legend.position = "none")
```


App made for our project, looks at the proportion of patients that are converted and demented given a range of values for range and normalized whole brain volume on the first visit. This app was designed for scalability in mind. As the population sample approaches n, these values will be more accurate and can be an accurate predictor of the probability of getting dementia.

```{r}

ui <- dashboardPage(skin = "red", 
                    dashboardHeader(title = "Predictors of Alzheimer's Disease From First Visit", 
                                    titleWidth = 600), 
                    dashboardSidebar(disable=T), 
                    dashboardBody(
  fluidPage(
  box(title = "Plot Options", width = 3,
  selectInput("nWBVMin", "Lower bound of normalized brain volume on first visit", choices=c("0.6","0.72","0.74","0.76","0.78","0.8","0.82")),
  selectInput("nWBVMax", "Upper bound of normalized brain volume on first visit",    choices=c("0.7","0.72","0.74","0.76","0.78","0.8","0.82","0.84")),
  selectInput("AgeMin", "Lower bound of Age", choices = c("60", "70", "80")),
  selectInput("AgeMax", "Upper bound of Age", choices = c("60", "70", "80", "90")),
   hr(),
      helpText("Reference:",
      "Marcus DS, Fotenos AF, Csernansky JG, Morris JC, Buckner RL. Open Access Series of Imaging Studies (OASIS): Longitudinal MRI Data in Nondemented and Demented Older Adults. Journal of cognitive neuroscience. 2010; 22(12):2677-2684. doi:10.1162/jocn.2009.21407.")
),
  box(width = 5, textOutput("result")),
  
  box(width = 5, title = "Number of Converted/Demented Patients", plotOutput("plot", width = "450px", height = "400px"))
)
)
)

server <- function(input, output, session) {
  
  totalDemented <- longitudinal %>%
    filter(group=='Demented') %>%
    nrow()
  
  totalConverted <- longitudinal %>%
    filter(group=='Converted') %>%
    nrow()
  #This gets the number of patients who fit the criteria
  numpatients <-  reactive({longitudinal %>%
      filter(visit==1) %>%
      filter(n_wbv>=input$nWBVMin & n_wbv<= input$nWBVMax) %>%
      filter(age>=input$AgeMin & age<= input$AgeMax)%>%
      nrow()
  })
  
  #This gets the number of patients who fit the criteria and are converted
  numconverted <-  reactive({longitudinal %>%
      filter(visit==1) %>%
      filter(n_wbv>=input$nWBVMin & n_wbv<= input$nWBVMax) %>%
      filter(age>=input$AgeMin & age<= input$AgeMax) %>%
      filter(group == 'Converted') %>%
      nrow()
  })
  
  #This gets the number of patients who fit the criteria and are demented
  numdemented <-  reactive({longitudinal %>%
      filter(visit==1) %>%
      filter(n_wbv>=input$nWBVMin & n_wbv<= input$nWBVMax) %>%
      filter(age>=input$AgeMin & age<= input$AgeMax) %>%
      filter(group == 'Demented') %>%
      nrow()
  })
  
  #Prints out the output
  output$result <- renderText({
    c(numpatients(),"PATIENTS MEET LISTED SEARCH CRITERIA: ", numdemented(), "of the 64 patients with existing dementia had brain volumes and ages within these values. ", numconverted(), "of the 14 who did not have dementia but later converted to dementia had brain volumes and ages within these values.")
  
    
    
    })
  
  output$plot <- renderPlot({
    longitudinal%>%
    filter(visit==1) %>%
    filter(n_wbv>=input$nWBVMin & n_wbv<= input$nWBVMax) %>%
    filter(age>=input$AgeMin & age<= input$AgeMax) %>%
    filter(group != 'Nondemented') %>%
    ggplot(aes(x = group)) + geom_bar(fill = 'red', alpha = 0.65) + ylim(0, 75) + geom_text(stat='count', aes(label=..count..), vjust=-1) + labs(x = 'Group', y = 'Count') 
    
  })
  
  # stop the app when closed
  session$onSessionEnded(stopApp)
}


shinyApp(ui, server)

```

### Conclusions

There are differences in several key variables such as Socioeconomic Status, Normalized Whole Brain Volume, and Average Educational Attainment between Demented and Non-Demented Patient Groups. Furthermore, We found that converted patients had a similar rate of decline in Normalized Whole Brain Volume  to Demented Patients, suggesting the possible predictive value of rate of change of Normalized Whole Brain values over time.



